---
title: "Timing experiments for dtwclust"
author: "Alexis Sarda-Espinosa"
output:
    html_vignette:
        toc: true
        number_sections: true
        fig_width: 7
        fig_height: 7
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignettePackage{dtwclust}
    %\VignetteIndexEntry{Timing experiments for dtwclust}
    %\VignetteEncoding{UTF-8}
bibliography: REFERENCES.bib
---

```{r setup, include=FALSE}
library("dtwclust")

load("../timing-experiments/dist-results.RData")
load("../timing-experiments/cent-results.RData")
load("../timing-experiments/tadpole-results.RData")
load("../timing-experiments/partitional-results.RData")

# knitr defaults
knitr::opts_chunk$set(echo = FALSE, comment = "#>")
```

# Introduction

To get an idea of the running times that can be expected for some of the included functions,
a series of timing experiments were made.
The data used again comes from the Character Trajectories set [@lichman2013].

These were run using `R` v3.4.1 and `dtwclust` v4.0.3.
The computer used was running GNU-Linux with an `i5-6500` Intel processor and 16GB of RAM.

# Distance experiments

## Calculations for single time-series

Results of the timing experiments for single time-series. 
An `NA` window size means that no constraint was used.
Note that the vertical scale is different for each facet.

```{r dist-single-plot, fig.height=7}
ggplot(dist_single_results,
       aes(x = factor(series_length),
           y = median_time_us,
           group = factor(window_size),
           colour = factor(window_size))) +
    geom_line() +
    facet_wrap(~distance, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

## Calculations for several time-series

### Lower bounds

First the results for the lower bounds.
The number of series in `y` is important because the envelopes are calculated for those.
In the case of `dtw_lb`, it means that there are more neighbors too.

```{r dist-multiple-lb-plot, fig.cap="The facets' columns indicate the number of parallel workers, whereas the rows indicate the distance that was used."}
id_gg <- grepl("lb", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = num_y,
           shape = factor(series_length))) +
    geom_point(size = 3) +
    facet_grid(distance ~ num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Shape-based distance

```{r dist-multiple-sbd-plot, fig.height=5, fig.cap="The facets' columns indicate the number of parallel workers."}
id_gg <- grepl("^sbd$", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = factor(series_length))) +
    geom_point(size = 3) +
    facet_wrap(~num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Dynamic time warping

```{r dist-multiple-dtw-plot, fig.height=8, fig.cap="The facets' columns indicate the number of parallel workers, whereas the rows indicate the window size that was used."}
id_gg <- grepl("^dtw_[um]", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = factor(series_length),
           shape = distance)) +
    geom_point(size = 3) +
    scale_shape_manual(values = c(0, 3)) +
    facet_grid(factor(window_size) ~ num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Triangular global alignment kernel

```{r dist-multiple-gak-plot, fig.height=8, fig.cap="The facets' columns indicate the number of parallel workers, whereas the rows indicate the window size that was used."}
id_gg <- grepl("^gak_", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = factor(series_length),
           shape = distance)) +
    geom_point(size = 3) +
    scale_shape_manual(values = c(0, 3)) +
    facet_grid(factor(window_size) ~ num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

# Prototyping experiments

## Shape extraction

```{r cent-shape-plot, fig.height=5}
id_gg <- grepl("^shape_", cent_results$cent)
ggplot(cent_results[id_gg,],
       aes(x = num_series,
           y = median_time_ms,
           colour = factor(series_length))) +
    geom_line() +
    facet_wrap(~cent) +
    theme_bw() + 
    theme(legend.position = "bottom")
```

## DTW barycenter averaging

```{r cent-dba-plot, fig.height=8, fig.cap="The facets' rows indicate the length of the considered series. In the columns, 'byS' means 'by series' and 'byV' means 'by variable' (see DBA documentation)."}
id_gg <- grepl("^dba_", cent_results$cent)
ggplot(cent_results[id_gg,],
       aes(x = num_series,
           y = median_time_ms,
           colour = factor(window_size))) +
    geom_line() +
    facet_grid(factor(series_length) ~ cent) +
    theme_bw() + 
    theme(legend.position = "bottom")
```

# Clustering experiments

With `tsclust`.

## TADPole

No parallel here because it is only parallelized with multiple `dc` values,
and that wasn't tested.

```{r clust-tadpole-plot}
ggplot(clus_tadpole_results,
       aes(x = num_series,
           y = median_time_s,
           colour = factor(window_size))) +
    geom_line() +
    facet_wrap(~lb) +
    theme_bw() + 
    theme(legend.position = "bottom")
```

## DTW special cases

### PAM centroids

The different flavors of PAM centroids.
The remarks about sparse matrices apply for other distances too.

```{r clust-part-dtw-pam-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$dtwlb_vs_dtwbasic$pam, 
                       id.vars=c("num_series", "k", "window_size"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = factor(window_size))) +
    geom_line() +
    facet_wrap(~variable) +
    theme_bw() +
    theme(legend.position = "bottom")
```

```{r clust-part-dtw-pam-reps-plot, fig.height=5}
cols <- setdiff(colnames(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps),
                "sparse_distmat_filled_percent")
ggdf <- reshape2::melt(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps[,cols], 
                       id.vars=c("num_series", "k", "num_repetitions"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = factor(num_repetitions))) +
    geom_line() +
    facet_wrap(~variable) +
    theme_bw() +
    theme(legend.position = "bottom")
```

```{r clust-part-dtw-pam-reps-distmat-plot, fig.height=5}
ggplot(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps,
       aes(x = num_series,
           y = sparse_distmat_filled_percent,
           colour = factor(num_repetitions))) +
    geom_line() +
    theme_bw() +
    theme(legend.position = "bottom")
```

### DBA centroids

This probably applies to other non-PAM centroids too.

```{r clust-part-dtw-dba-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$dtwlb_vs_dtwbasic$dba, 
                       id.vars=c("num_series", "k", "window_size"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = factor(window_size))) +
    geom_line() +
    facet_wrap(~variable) +
    theme_bw() +
    theme(legend.position = "bottom")
```

## Effect of `k` on PAM centroids with sparse matrices

Test different `k` values.
Using non-symmetric DTW and symmetric SBD.

### Non-symmetric

```{r clust-part-sparse-pam-k-nonsymmetric-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$sparse_pam_k$non_symmetric, 
                       id.vars=c("num_series", "k"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = factor(k))) +
    geom_line() +
    facet_wrap(~variable, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Symmetric

```{r clust-part-sparse-pam-k-symmetric-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$sparse_pam_k$symmetric, 
                       id.vars=c("num_series", "k"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = factor(k))) +
    geom_line() +
    facet_wrap(~variable, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Remarks

The effect of `k` is small,
and as expected smaller values fill the matrix faster.
But if we consider the results with many repetitions above,
then it doesn't make much sense to use sparse matrices.

# References
