---
title: "Timing experiments for dtwclust"
author: "Alexis Sarda-Espinosa"
output:
    html_vignette:
        toc: true
        number_sections: true
        fig_width: 7
        fig_height: 7
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignettePackage{dtwclust}
    %\VignetteIndexEntry{Timing experiments for dtwclust}
    %\VignetteEncoding{UTF-8}
bibliography: REFERENCES.bib
---

```{r setup, include=FALSE}
library("dtwclust")
data("dtwclustTimings")

dist_single_results <- dtwclustTimings$dist$single
dist_multiple_results <- dtwclustTimings$dist$multiple
cent_results <- dtwclustTimings$cent
clus_tadpole_results <- dtwclustTimings$tadpole
partitional_results <- dtwclustTimings$partitional

factor_columns <- c("series_length", "window_size", "k", "num_repetitions")
adjust_factors <- function(df) {
    for (column in factor_columns) {
        if (column %in% colnames(df)) {
            df[[column]] <- factor(df[[column]])
        }
    }
    df
}

dist_single_results <- adjust_factors(dist_single_results)
dist_multiple_results <- adjust_factors(dist_multiple_results)
cent_results <- adjust_factors(cent_results)
clus_tadpole_results <- adjust_factors(clus_tadpole_results)
partitional_results$dtwlb_vs_dtwbasic$pam <- adjust_factors(partitional_results$dtwlb_vs_dtwbasic$pam)
partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps <- adjust_factors(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps)
partitional_results$dtwlb_vs_dtwbasic$dba <- adjust_factors(partitional_results$dtwlb_vs_dtwbasic$dba)
partitional_results$sparse_pam_k$non_symmetric <- adjust_factors(partitional_results$sparse_pam_k$non_symmetric)
partitional_results$sparse_pam_k$symmetric <- adjust_factors(partitional_results$sparse_pam_k$symmetric)

# knitr defaults
knitr::opts_chunk$set(echo = FALSE, comment = "#>")
```

# Introduction

To get an idea of the running times that can be expected for some of the included functions,
a series of timing experiments were made.
The data used again comes from the Character Trajectories set [@lichman2013].

These were run using `R` v3.4.1 and `dtwclust` v4.0.3.
The computer used was running GNU/Linux (LTS kernel v4.9) with an `i5-6500` Intel processor (4 cores) and 16GB of RAM.
They took almost 58 hours to complete (208183.266 seconds).

# Distance experiments

## Calculations for single time-series

Results of the timing experiments for single time-series. 
An `NA` window size means that no constraint was used.
Note that the vertical scale is different for each facet.

```{r dist-single-plot, fig.height=7, fig.cap="*Note the different vertical scales for each facet, even though they are all in microseconds.*"}
ggplot(dist_single_results,
       aes(x = series_length,
           y = median_time_us,
           group = window_size,
           colour = window_size)) +
    geom_line() +
    facet_wrap(~distance, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

## Calculations for several time-series

### Lower bounds

First the results for the lower bounds.
The number of series in `y` is important because the envelopes are calculated for those.
In the case of `dtw_lb`, it means that there are more neighbors too.

```{r dist-multiple-lb-plot, fig.cap="*The facets' columns indicate the number of parallel workers, whereas the rows indicate the distance that was used. The vertical scales are different for each row, but they are all in milliseconds.*"}
id_gg <- grepl("lb", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = num_y,
           shape = series_length)) +
    geom_point(size = 3) +
    facet_grid(distance ~ num_workers, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Shape-based distance

```{r dist-multiple-sbd-plot, fig.height=5, fig.cap="*The facets' columns indicate the number of parallel workers.*"}
id_gg <- grepl("^sbd$", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = series_length)) +
    geom_point(size = 3) +
    facet_wrap(~num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Dynamic time warping

```{r dist-multiple-dtw-plot, fig.height=8, fig.cap="*The facets' columns indicate the number of parallel workers, whereas the rows indicate the length of the series being considered (all series had the same length for each case). All times are in milliseconds.*"}
id_gg <- grepl("^dtw_[um]", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = window_size,
           shape = distance)) +
    geom_point(size = 3) +
    scale_shape_manual(values = c(0, 3)) +
    facet_grid(series_length ~ num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Triangular global alignment kernel

```{r dist-multiple-gak-plot, fig.height=8, fig.cap="*The facets' columns indicate the number of parallel workers, whereas the rows indicate the length of the series being considered (all series had the same length for each case). All times are in milliseconds.*"}
id_gg <- grepl("^gak_", dist_multiple_results$distance)
ggplot(dist_multiple_results[id_gg,],
       aes(x = num_total,
           y = median_time_ms,
           colour = window_size)) +
    geom_point(size = 3) +
    facet_grid(series_length ~ num_workers) +
    theme_bw() +
    theme(legend.position = "bottom")
```

# Prototyping experiments

## Shape extraction

```{r cent-shape-plot, fig.height=5}
id_gg <- grepl("^shape_", cent_results$cent)
ggplot(cent_results[id_gg,],
       aes(x = num_series,
           y = median_time_ms,
           colour = series_length)) +
    geom_line() +
    facet_wrap(~cent) +
    theme_bw() + 
    theme(legend.position = "bottom")
```

## DTW barycenter averaging

```{r cent-dba-plot, fig.height=8, fig.cap="*The facets' rows indicate the length of the considered series. In the columns, 'byS' means 'by series' and 'byV' means 'by variable' (see DBA documentation).*"}
id_gg <- grepl("^dba_", cent_results$cent)
ggplot(cent_results[id_gg,],
       aes(x = num_series,
           y = median_time_ms,
           colour = window_size)) +
    geom_line() +
    facet_grid(series_length ~ cent) +
    theme_bw() + 
    theme(legend.position = "bottom")
```

# Clustering experiments

With `tsclust`.

## TADPole

No parallel here because it is only parallelized with multiple `dc` values,
and that wasn't tested.

```{r clust-tadpole-plot, fig.cap="*In the facets, lbk stands for lb_keogh and lbi for lb_improved.*"}
ggplot(clus_tadpole_results,
       aes(x = num_series,
           y = median_time_s,
           colour = window_size)) +
    geom_line() +
    facet_wrap(~lb) +
    theme_bw() + 
    theme(legend.position = "bottom")
```

## DTW special cases

### PAM centroids

The different flavors of PAM centroids.
The remarks about sparse matrices apply for other distances too.

```{r clust-part-dtw-pam-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$dtwlb_vs_dtwbasic$pam, 
                       id.vars=c("num_series", "k", "window_size"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = window_size)) +
    geom_line() +
    facet_wrap(~variable) +
    theme_bw() +
    theme(legend.position = "bottom")
```

```{r clust-part-dtw-pam-reps-plot, fig.height=5}
cols <- setdiff(colnames(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps),
                "sparse_distmat_filled_percent")
ggdf <- reshape2::melt(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps[,cols], 
                       id.vars=c("num_series", "k", "num_repetitions"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = num_repetitions)) +
    geom_line() +
    facet_wrap(~variable) +
    theme_bw() +
    theme(legend.position = "bottom")
```

```{r clust-part-dtw-pam-reps-distmat-plot, fig.height=5}
ggplot(partitional_results$dtwlb_vs_dtwbasic$pam_vs_reps,
       aes(x = num_series,
           y = sparse_distmat_filled_percent,
           colour = num_repetitions)) +
    geom_line() +
    theme_bw() +
    theme(legend.position = "bottom")
```

### DBA centroids

This probably applies to other non-PAM centroids too.

```{r clust-part-dtw-dba-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$dtwlb_vs_dtwbasic$dba, 
                       id.vars=c("num_series", "k", "window_size"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = window_size)) +
    geom_line() +
    facet_wrap(~variable) +
    theme_bw() +
    theme(legend.position = "bottom")
```

## Effect of `k` on PAM centroids with sparse matrices

Test different `k` values.
Using non-symmetric DTW and symmetric SBD.

### Non-symmetric

```{r clust-part-sparse-pam-k-nonsymmetric-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$sparse_pam_k$non_symmetric, 
                       id.vars=c("num_series", "k"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = k)) +
    geom_line() +
    facet_wrap(~variable, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Symmetric

```{r clust-part-sparse-pam-k-symmetric-plot, fig.height=5}
ggdf <- reshape2::melt(partitional_results$sparse_pam_k$symmetric, 
                       id.vars=c("num_series", "k"))
ggplot(ggdf,
       aes(x = num_series,
           y = value,
           colour = k)) +
    geom_line() +
    facet_wrap(~variable, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "bottom")
```

### Remarks

The effect of `k` is small,
and as expected smaller values fill the matrix faster.
But if we consider the results with many repetitions above,
then it doesn't make much sense to use sparse matrices.

# References
